---
title: "Access iNaturalist data through APIs"
author: "Edmund Hart, Stéphane Guillou"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Access iNaturalist data through APIs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE)
```

## About

rinat is a wrapper for iNaturalist APIs for accessing the observations. The detailed documentation of the API is available on the [iNaturalist website](https://www.inaturalist.org/pages/api+reference) and is part of our larger species occurrence searching packages [SPOCC](https://github.com/ropensci/spocc).


## Quickstart guide

### Get observations

`get_inat_obs()` is the primary function that retrieves observations from iNaturalist. The text or taxon search can be refined by observation date, record quality and location.

> It is recommended to set the `quality` argument to `"research"` in order to get more reliable data that has been validated by several contributors.

#### Fuzzy search

You can search for observations by either common or scientific name. It will search the entire iNaturalist database, so the search below will return entries that _mention_ Monarch butterflies, not just Monarch observations.

```{r}
library(rinat)
monarchs <- get_inat_obs(query = "Monarch Butterfly", year = 2021)
unique(monarchs$scientific_name)
```

> Note that `get_inat_obs()` will return 100 observations by default. This can be controlled with the `maxresults` argument.

Another use for a fuzzy search is searching for a habitat, e.g. searching for all observations that might happen in a [vernal pool](https://en.wikipedia.org/wiki/Vernal_pool). We can then explore the taxon names found.

```{r}
vp_obs <- get_inat_obs(query = "vernal pool")
# see the first few taxa
head(unique(vp_obs$scientific_name))
```


#### Taxon search

To return only records of a specific species or taxonomic group, use the `taxon_name` argument. For example, to return observations of anything from the Nymphalidae family, and restricting the search to the year 2015:

```{r}
nymphalidae <- get_inat_obs(taxon_name  = "Nymphalidae", year = 2015)
# how many unique taxa?
length(unique(nymphalidae$scientific_name))
```

And to return only the Monarch butterfly observations that also mention the term "chrysalis":

```{r}
monarch_chrysalis <- get_inat_obs(taxon_name = "Danaus plexippus", query = "chrysalis")
```


#### Bounding box search

You can also search within a bounding box by giving a simple set of coordinates.

```{r fig.width=7, fig.height=4, fig.retina=3}
## Search by area
bounds <- c(38.44047, -125, 40.86652, -121.837)
deer <- get_inat_obs(query = "Mule Deer", bounds = bounds)
plot(deer$longitude, deer$latitude)
```


### Other functions

#### Get information and observations by project

You can get all the observations for a project if you know its ID or name as an iNaturalist slug.

```{r}
## Just get info about a project
vt_crows <- get_inat_obs_project("crows-in-vermont", type = "info", raw = FALSE)
```

```{r}
## Now get all the observations for that project
vt_crows_obs <- get_inat_obs_project(vt_crows$id, type = "observations")
```


#### Get observation details

Detailed information about a specific observation can be retrieved by observation ID. The easiest way to get the ID is from a previous search for that taxon.

```{r}
head(get_inat_obs_id(monarch_chrysalis$id[1]))
```


#### Get all observations by user

If you just want all the observations by a user you can download all their observations by user ID. A word of warning though, this can be quite large (easily into the 1000's).

```{r}
user_obs <- get_inat_obs_user(monarch_chrysalis$user_login[1], maxresults = 20)
head(user_obs)[,1:5]
```


#### Stats by taxa

Basic statistics are available for taxa counts by date, date range, place ID (numeric ID), or user ID (string). Only the top 5 species are listed.

```{r}
## By date
counts <- get_inat_taxon_stats(date = "2020-06-14")
counts$total
### Top 5 species
counts$species_counts
### Most common taxon ranks
counts$rank_counts
```


#### Stats by user

Similar statistics can be gotten for users. The same input parameters can be used.

```{r}
## By date
counts <- get_inat_user_stats(date = "2010-06-14")
counts$total
counts$most_observations[1:10,]
counts$most_species[1:10,]
```

```{r}
## By place_ID
vt_crows <- get_inat_obs_project("crows-in-vermont", type = "info", raw = FALSE)
place_counts <- get_inat_user_stats(place = vt_crows$place_id)
place_counts$total
place_counts$most_observations[1:10,]
place_counts$most_species[1:10,]
```


### Mapping

Basic maps can be created with the `inat_map()` function to quickly visualize search results. The `plot = FALSE` option can be used to avoid displaying the initial plot when further customising it with ggplot2 functions.

```{r fig.width=7, fig.height=4, fig.retina=3}
library(ggplot2)

## Map 100 spotted salamanders
a_mac <- get_inat_obs(taxon_name = "Ambystoma maculatum", year = 2021)
salamander_map <- inat_map(a_mac, plot = FALSE)

### Further customise the returned ggplot object
salamander_map + borders("state") + theme_bw()
```

`inat_map()` is useful for quickly mapping single-species data obtained with rinat. However, more complicated plots are best made from scratch. Here is an example of customised map that does not make use of it. (Note the use of `quality = "research"` to restrict the search to the more reliable observations.)

```{r fig.width=7, fig.height=7, fig.retina=3}
## A more elaborate map of Colibri sp.
colibri <- get_inat_obs(taxon_name = "Colibri",
                        quality = "research",
                        maxresults = 500)
ggplot(data = colibri, aes(x = longitude,
                         y = latitude,
                         colour = scientific_name)) +
  geom_polygon(data = map_data("world"),
                   aes(x = long, y = lat, group = group),
                   fill = "grey95",
                   color = "gray40",
                   size = 0.1) +
  geom_point(size = 0.7, alpha = 0.5) +
  coord_fixed(xlim = range(colibri$longitude, na.rm = TRUE),
              ylim = range(colibri$latitude, na.rm = TRUE)) +
  theme_bw()
```

