
---
title: "Access iNaturalist data through APIs"
author: "Edmund Hart"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Access iNaturalist data through APIs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Quickstart guide
 
## About

R wrapper for iNaturalist APIs for accessing the observations. The Detailed documentation of API is available on the [iNaturalist website](http://www.inaturalist.org/pages/api+reference) and is part of our larger species occurrence searching packages [SPOCC](http://github.com/ropensci/spocc)

## Get observations

__Searching__

_Fuzzy search_

You can search for observations by either common or Latin name. It will search the entire iNaturalist entry, so the search below will return all entries that mention Monarch butterflies, not just entries for Monarchs.

```{r,  message=FALSE,warning=FALSE,echo=FALSE}
options(stringsAsFactors = FALSE)
```

```{r, message=FALSE,warning=FALSE}
library(rinat)

butterflies <- get_inat_obs(query = "Monarch Butterfly")
```

Another use for a fuzzy search is searching for a common name or habitat, e.g. searching for all observations that might happen in a vernal pool. We can then see all the species names found.

```{r, message=FALSE,warning=FALSE}
vp_obs <- get_inat_obs(query = "vernal pool")
head(vp_obs$species_guess)
```

_Taxon query_

To return only records for a specific species or taxonomic group, use the taxon option.


```{r}
## Return just observations in the family Plecoptera
stone_flies <- get_inat_obs(taxon_name  = "Plecoptera", year = 2010)

## Return just Monarch Butterfly records
just_butterflies <- get_inat_obs(taxon_name = "Danaus plexippus")
```

_Bounding box search_

You can also search within a bounding box by giving a simple set of coordinates.

```{r}
## Search by area
bounds <- c(38.44047, -125, 40.86652, -121.837)
deer <- get_inat_obs(query = "Mule Deer", bounds = bounds)
```

__Other functions__

_Get information and observations by project_

You can get all the observations for a project if you know it's ID or name as an iNaturalist slug:

```{r}
## Just get info about a project
vt_crows <- get_inat_obs_project("crows-in-vermont", type = "info", raw = FALSE)
```

```{r}
## Now get all the observations for that project
vt_crows_obs <- get_inat_obs_project(vt_crows$id, type = "observations")
```

_Get observation details_

Detailed information about a specific observation can be retrieved by observation ID. The easiest way to get the ID is from a more general search.

```{r}
m_obs <- get_inat_obs(query = "Monarch Butterfly")
head(get_inat_obs_id(m_obs$id[1]))
```

_Get all observations by user_

If you just want all the observations by a user you can download all their observations by user ID. A word of warning though: this can be quite large (easily into the 1000's).

```{r}
m_obs <- get_inat_obs(query = "Monarch Butterfly")
head(get_inat_obs_user(as.character(m_obs$user_login[1]), maxresults = 20))[,1:5]
```

_Stats by taxa_

Basic statistics are available for taxa counts by date, date range, place ID (numeric ID), or user ID (string):

```{r}
## By date
counts <- get_inat_taxon_stats(date = "2010-06-14")
print(counts$total)
print(counts$species_counts[1:5,])
print(counts$rank_counts)
```

_Stats by user_

Similar statistics can be gotten for users. The same input parameters can be used, but results are the top five users by species count and observation count.

```{r}
## By date
counts <- get_inat_user_stats(date = "2010-06-14")
print(counts$total)
print(counts$most_observations[1:10,])
print(counts$most_species[1:10,])
```

```{r}
## By place_ID
vt_crows <- get_inat_obs_project("crows-in-vermont", type = "info", raw = FALSE)
```

```{r}
place_counts <- get_inat_user_stats(place = vt_crows$place_id)
print(place_counts$total)
print(place_counts$most_observations[1:10,])
print(place_counts$most_species[1:10,])
```

## Mapping

Basic maps can be created as well to quickly visualize search results. Maps can either be plotted automatically with `plot = TRUE` or simply return a ggplot2 object with `plot = FALSE`. This works well with single species data, but more complicated plots are best made from scratch.


```{r fig.width=7,fig.height=4}
library(ggplot2)

## Map salamanders in the genus Ambystoma
m_obs <- get_inat_obs(taxon_name = "Ambystoma maculatum")

## Return a ggplot2 object instead...
salamander_map <- inat_map(m_obs, plot = FALSE)
## ...so we can modify the returned map further
salamander_map + borders("state") + theme_bw()
```




